smdi: An R package to perform structural missing data investigations on partially observed confounders in real-world databases

### 2023-08-03 manuscript/manuscript.qmd
stopped at paragraph about set visualization
finalized main body text draft; discussion next

```{r setup}
#| label: setup
#| include: false
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(gtsummary)
library(flextable)
library(officer)
library(captioner)

@@

::: {.callout-note icon="false"}
## Highlights

@@

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and is consequently not often performed in studies [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostic analyses to scrutinize partially observed confounder data which had not been implemented elsewhere yet and complement other analytic R packages regularly used in RWE studies.

# Methods

## **Package details and installation instructions**

The `smdi` R package was written in R language (version 4.2.1) and using the devtools [@devtools] and roxygen [@roxygen2] packages. The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command. To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis.

@@

@fig-workflow illustrates the recommended workflow and functions that can be used to systematically approach diagnostics on partially observed confounders.

In general, the workflow comprises two broader diagnostic categories: descriptives and pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. This section covers the basic ideas behind the main package functions, the parameters a user can specify, the results that are being returned by calling the functions and the interpretation of results. Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unqiue patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions consider automatically any variable in the dataframe that exhibits at least one missing value.

@@

As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates. The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations for partially observed covariates which can also be stratified by one or more grouping variables (e.g., by an exposure or outcome variable). The `smdi_vis()` function returns a corresponding bar chart plot.

[@ruddle2022]

@@

The main functions used to investigate the potential underlying missing data mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework on inference and missing data [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with corresponding minimum and maximum AMSDs observed. The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization of this, the function also creates a *ggplot2* graph [@ggplot2] illustrating the ASMDs for each compared patient characteristic in descending order.

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this is a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate individually. In contrast, Little's test [@little1988; @naniar] (`smdi_little()`) computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` functions. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to potential MCAR or MNAR mechanisms.

The function returns an object of class `rf` which generically displays an overview of the AUC values of all partially observed covariates. The AUC values are based on the prediction made in the respective test datasets. For more details, the `rf` object additionally includes a *ggplot2* graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy. This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. For tuning the random forest model, the `smdi_rf()` function has a few parameters that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.

#### Group 3 Diagnostics

@@

@@

@@

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "workflow.jpg"))
```
:::

>>>>

```{r setup}
#| label: setup
#| include: false
library(cowplot)
library(here)
library(tidyverse)
library(readxl)
library(gt)
library(gtsummary)
library(flextable)
library(officer)
library(captioner)
library(smdi)
suppressPackageStartupMessages(library(ggimage))

@@

::: {.callout-tip icon="false"}
## Lay Summary

@@

In order to make an informed decision about the most appropriate analytic approach to arrive at unbiased estimates, it is crucial to investigate and understand the potential patterns and mechanisms that underlie the partially observed confounder data [@vanbuuren2018; @rubin1976; @little2019]. Usually these are not known for a given RWE study but general guidance papers and frameworks have suggested various routine diagnostics to investigate missing data patterns and mechanisms (see definitions box). These methods comprise standard procedures such as comparing baseline characteristics and outcomes between patients with and without the partially observed covariate(s) [@lee2021; @sondhi2023a; @hotelling1931; @little1988; @pedersen2017], checking the ability to predict missingness based on observed data [@sondhi2023a] and assessing if causal relationships between variables and their missingness are recoverable based on available data [@madley-dowd2019] guided by directed acyclic graphs [@Lee2023; @Moreno-Betancur2018] or M-graphs [@mohan2021]. However, these methods have so far been only described and tested in isolation from each other and no principled approach exists. In addition, the practical implementation of all of these diagnostics is time-consuming, tedious and not often performed [@carroll2020; @wood2004; @harel2012].

@@

To streamline and ease the implementation of these routine missing data diagnostics for confounder data in RWE studies, we developed the `smdi` (structural missing data investigations) R package [@smdi]. The goal of `smdi` is to provide a toolbox of different diagnostics to analyze the potentially underlying missingness patterns and mechanisms of partially observed confounders in RWE studies.

# Materials and Methods

The `smdi` R package was written in R language (version 4.2.1) employing the devtools [@devtools] and roxygen [@roxygen2] packages. The package is available on CRAN (<https://cran.r-project.org/web/packages/smdi)>, Posit Public Package Manager (<https://packagemanager.posit.co/client/#/repos/2/packages/smdi>) and GitLab (<https://gitlab-scm.partners.org/janickweberpals/smdi>) and can be installed via the `install.packages("smdi")` command. To ensure the quality and robustness of the package and corresponding functions, we implemented comprehensive unit tests with a coverage of 95.81% and established automated R CMD checks [@wickham2023r] via continuous integration and deployment which are scheduled to execute on a daily basis. Additional resources such as documentation, articles and vignettes are provided on the package website under <https://janickweberpals.gitlab-pages.partners.org/smdi>.

@@

@fig-workflow illustrates the recommended workflow and functions to systematically approach diagnostics on partially observed confounders.

In general, the workflow comprises two broader diagnostic categories: descriptives/pattern diagnostics and inferential diagnostics on potentially underlying missingness mechanisms. In this section, we cover the principles behind the main package functions, the parameters a user can specify, the returned results and the interpretation of results. Examples are illustrated using a simulated EHR dataset that is part of the package (more details can be found under <https://janickweberpals.gitlab-pages.partners.org/smdi/articles/a_data_generation.html>). Generally, for all functions in the `smdi` package, a *dataframe* is expected (`data` parameter) as input with a format where one row represents one unique patient and the columns stand for the variables relevant for the study, i.e., the exposure, outcome, fully observed covariates and the partially observed covariates. Any non-informative columns, such as patient identifiers, date columns or zip codes should be dropped from the dataframe before calling the functions. Throughout all functions, the user has the option to specify a vector with the column name(s) of the partially observed covariate(s) that should be investigated (`covar` parameter). If nothing is specified, all functions automatically consider any variable in the dataframe that exhibits at least one missing value.

@@

As a first step to explore the missingness in new datasets, the `smdi` package provides a few basic functions to describe and summarize missingness across all covariates. The `smdi_summarize()` function takes the dataframe as input and returns the amount and proportion of missing observations, which can also be stratified by a grouping variable (e.g., by an exposure or outcome variable). The `smdi_vis()` function returns a corresponding bar chart plot (example @fig-examples a).

To visually inspect potential missing data patterns, we re-exported the `gg_miss_upset()` function of the naniar package [@naniar-2]. This function uses a set visualization technique to visually infer potential (non-)monotone patterns based on the number of intersecting missing observations across all partially observed covariates [@ruddle2022]. For example, a monotone pattern would be visually evident if, for a set of two or more variables (e.g., lab1, lab2), all or the majority of missing records would be observed for both labs simultaneously (example @fig-examples b).

@@

The core functions to infer potentially underlying missingness mechanisms are categorized into three group diagnostics based on their general analytic properties (**`r table_cap("tbl1", display = "cite")`**).

@@

The aim of the `smdi_asmd()`, `smdi_hotelling()` and `smdi_little()` functions is to explore dissimilarities in patient characteristics between those with and without observed values for the partially observed covariate. According to Rubin's framework [@rubin1976], when missingness is at random (MAR), it can be explained by observed covariates. Consequently, significant differences in patient characteristics would be expected under a MAR mechanism between strata of patients with and without the partially observed covariate. If the missingness depends only on unobserved factors (missing not at random \[MNAR\]) or does not depend on either observed or unobserved covariates (missing completely at random \[MCAR\]), differences should not be observable.

To quantify and test such differences, the `smdi_asmd()` function computes absolute standardized mean differences (ASMD) of observed patient characteristics [@schober2019; @austin2011; @tableone]. The function returns an *asmd* object which displays an aggregated summary of the average or median ASMDs (which can be flexibly chosen using the `median` parameter) along with a corresponding range of minimum and maximum AMSDs observed. The *asmd* object also returns a detailed 'Table 1' for each partially observed covariate displaying the distributions of observed patient characteristics and resulting ASMDs between patients with and without an observed value for the partially observed covariate. For a graphical visualization of this, the function returns a graph illustrating the ASMD for each compared patient characteristic (example @fig-examples c) [@ggplot2] .

The `smdi_hotelling()` and `smdi_little()` functions complement the `smdi_asmd()` function by examining the differences in patient characteristics as a formal statistical hypothesis test. Hotelling's test [@hotelling1931; @Hotelling] formalizes this as a multivariate t-test for each partially observed covariate, which means that `smdi_hotelling()` returns a test statistic and a corresponding p-value for each each partially observed covariate separately. In contrast, `smdi_little()` [@little1988; @naniar] computes a single global chi-square test statistic and p-value across all partially observed covariates with the null hypothesis that the data is MCAR.

@@

The group 2 diagnostics assesses the ability to predict missingness based on observed covariates via the `smdi_rf()` function. This function trains and fits a random forest classification model [@randomForest; @sondhi2023a] to predict the missing indicator of the partially observed covariate given observed covariates as the predictors. If the resulting area under the receiver operating characteristic curve (AUC) is meaningfully higher than 0.5, this would give some evidence for MAR being the underlying missingness mechanism. In case of values close 0.5, this would rather indicate a random prediction and translate to a potential MCAR or MNAR mechanism.

The function returns an object of class *rf* which generically prints an overview of the AUC values of all partially observed covariates. The AUC value is based on the prediction made in the respective test dataset. The *rf* object further returns a graph for each partially observed covariate displaying the relative importance of the predictors in the training dataset expressed as the mean decrease in accuracy (example @fig-examples d). This metric can be valuable for interpreting and identifying strong predictors of missingness. It quantifies how much the accuracy of the prediction (i.e., the ratio of correct predictions to the total number of predictions made) would decrease if we excluded a specific predictor from the prediction model. In case of inflated AUC values (\>0.9), the function prompts a message to the user detailing the most important predictor. If another partially observed covariate is identified as a perfect predictor, the presence of a monotone missing data pattern may be likely in which case it is recommended to exclude this covariate and run the diagnostics for each partially observed covariate independently.

```{=html}
<!--

To tune the random forest model hyperparameters, the `smdi_rf()` function has a few arguments that can be specified by the user for optimization such as the number of trees to grow (`ntree`; default is 1,000 trees), the ratio of the split between train and test datasets (`train_test_ratio`; default is a 70/30 split) and the number of cores to parallelize the computation on (`n_cores`, default is 1), since this function can be very time consuming, especially with larger datasets.

-->
```
#### Group 3 Diagnostics

@@

#### `smdi_diagnose()` to compute all three group diagnostics

The `smdi_diagnose()` function enables a user to compute all of the three group diagnostics introduced above with just one function call.

```{r}
#| eval: false
#| echo: true
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)"
  ) %>% 
  smdi_style_gt()
```

The function will return an object of class *smdi* containing a table with the results of all diagnostics for each specified partially observed covariate and the global Little's test p-value across all covariates (**`r table_cap("tbl2", display = "cite")`**). The `smdi_style_gt()` function is an ancillary function that takes an object of class *smdi* and produces a formatted and publication-ready gt table [@gt] which can be seamlessly exported to different file formats (e.g., .docx, .pdf, etc.) for reports or manuscripts.

@@

Missing data are ubiquitous in real-world databases and may lead to bias and spurious results if not handled appropriately.

By comparing the resulting diagnostic parameters to expected estimates, based on previous simulation results [@weberpals2023], the diagnostics can provide valuable insights into underlying missingness structures and help elucidate if analytic approaches such as missing imputation analyses are viable options @fig-results.

@@

`r table_cap(name = "tbl2", "Example output of the smdi_diagnose() function applied to the examplary smdi_data dataset.")`

```{r}
#| echo: false
smdi_diagnose(
  data = smdi_data,
  covar = NULL, # NULL includes all covariates with at least one NA
  model = "cox",
  form_lhs = "Surv(eventtime, status)",
  n_cores = 3,
  ) %>% 
  smdi_style_gt()
```

@@

```{r}
#| label: fig-workflow
#| fig-cap: "Overview of all `smdi` functions and suggested workflow to conduct missing data diagnostics. *gg_miss_upset() and md.pattern() are re-exports of the naniar and mice package, respectively."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "Figure_1_workflow.jpg"))
```

```{r}
#| label: fig-examples
#| fig-cap: "Exemplary visual outputs of the a) smdi_vis(), b) gg_miss_upset(), c) smdi_asmd() and d) smdi_rf() functions, respectively."
#| echo: false
#| warning: false
#| fig-dpi: 400
#| fig-width: 15
#| fig-height: 10
#| out-width: 110%

# a) smdi_vis()
panel_a <- smdi_vis(data = smdi_data, strata = "exposure")
panel_a <- panel_a + ggplot2::theme_bw(base_size = 15)

# b) gg_miss_upset()
# for panel B, we simulatea monotone missingness pattern
# following an MCAR mechanism
set.seed(42)

data_monotone <- smdi_data_complete %>% 
  mutate(
    lab1 = rnorm(nrow(smdi_data_complete), mean = 5, sd = 0.5),
    lab2 = rnorm(nrow(smdi_data_complete), mean = 10, sd = 2.25)
    )

data_monotone[3:503, "lab1"] <- NA
data_monotone[1:500, "lab2"] <- NA

panel_b <- as.ggplot(smdi::gg_miss_upset(data = data_monotone))

#smdi_asmd
asmd <- smdi_asmd(data = smdi_data)
panel_c <- asmd$pdl1_num$asmd_plot

#smdi_asmd
rf <- smdi_rf(data = smdi_data, n_cores = 3)
panel_d <- rf$ecog_cat$rf_plot

panel <- cowplot::plot_grid(
  panel_a, panel_b, panel_c, panel_d,
  ncol = 2,
  nrow = 2,
  labels = c("a)", # % missing
             "b)", # set visualization
             "c)", # asmd
             "d)" # rf
             ),
  label_size = 18,
  label_colour = "black",
  label_fontface = "bold",
  vjust = 1.25,
  hjust = -.25
  )

panel

jpeg(here::here("figures", "Figure_2_visuals.jpg"), width = 16, height = 10, units = "in", res=400)
panel
invisible(dev.off())
```
:::

```{r}
#| label: fig-results
#| eval: true
#| fig-cap: "Example of how smdi diagnostics can be applied to compute and compare diagnostic parameters of partially observed covariates to expected parameters of common missingness mechanisms based on a former large-scale simulation study [@weberpals2023]."
#| echo: false
#| out-width: 110%
knitr::include_graphics(here::here("figures", "Figure_3_results.jpg"))
```
